---
title: "permabrioche: Repeated-Measures PERMANOVA on Simulated Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{permabrioche (simulated example)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.align = "center"
)
set.seed(1)
```

```{r setup}
library(permabrioche)
library(vegan)
library(dplyr)
library(ggplot2)
```

# Overview

This vignette illustrates how to use **permabrioche** on a simple simulated
repeated-measures microbiome dataset. We:

1. simulate subjects with multiple samples and microbial taxa,  
2. construct a binary exposure that varies within subjects,  
3. compare:
   - a naïve PERMANOVA that ignores subject,
   - a repeated-measures PERMANOVA using `PERMANOVA_repeat_measures()`,
   - a Hájek-style test using `hajek_repeat_measures()`,
4. run a small simulation to inspect type I error (null, no effect) and power
   (alternative, spiked effect).

# Simulated repeated-measures microbiome data

We simulate:

- `n_subj` subjects,
- each with `m` repeated samples,
- `p` taxa per sample,
- log-normal abundances with a subject-specific random effect.

```{r simulate-once}
simulate_microbiome <- function(
  n_subj      = 20,
  m_per_subj  = 3,
  n_taxa      = 50,
  effect_size = 0,     # 0 = null, >0 = spike effect on exposure = 1
  spike_frac  = 0.1    # fraction of taxa to spike with effect
) {
  # subjects and samples
  subj_ids <- paste0("S", seq_len(n_subj))
  meta <- do.call(rbind, lapply(subj_ids, function(s) {
    data.frame(
      subject = s,
      sample  = paste0(s, "_T", seq_len(m_per_subj)),
      stringsAsFactors = FALSE
    )
  }))

  # subject-level random effects on log-scale
  subj_effects <- rnorm(n_subj, mean = 0, sd = 1)
  names(subj_effects) <- subj_ids

  # base log-abundances per taxon
  taxon_means <- rnorm(n_taxa, mean = 0, sd = 1)
  tax_names   <- paste0("Taxon", seq_len(n_taxa))

  # construct matrix (samples × taxa)
  X <- matrix(NA_real_, nrow = nrow(meta), ncol = n_taxa,
              dimnames = list(meta$sample, tax_names))

  for (i in seq_len(nrow(meta))) {
    s_id <- meta$subject[i]
    subj_shift <- subj_effects[s_id]
    X[i, ] <- rnorm(n_taxa, mean = taxon_means + subj_shift, sd = 0.5)
  }

  # convert to positive abundances (log-normal-ish)
  X <- exp(X)

  # within-subject binary exposure
  meta <- meta %>%
    group_by(subject) %>%
    mutate(exposure = sample(c(0L, 1L), n(), replace = TRUE)) %>%
    ungroup()

  # spike effect if effect_size > 0
  if (effect_size > 0) {
    n_spike <- max(1L, round(spike_frac * n_taxa))
    spike_taxa <- sample(tax_names, n_spike)

    # multiply abundances for exposure = 1 by (1 + effect_size)
    X[meta$exposure == 1, spike_taxa] <-
      X[meta$exposure == 1, spike_taxa] * (1 + effect_size)
  }

  # renormalize rows to relative abundances
  X <- sweep(X, 1, rowSums(X), FUN = "/")

  list(
    metadata = meta,
    bugs     = X
  )
}

# Quick check
sim0 <- simulate_microbiome()
str(sim0$metadata)
dim(sim0$bugs)
```

# One-shot analysis: null scenario

We start with a **null** scenario, where `effect_size = 0` (no true effect of
`exposure` on the microbiome).

```{r one-null}
sim_null <- simulate_microbiome(effect_size = 0)

meta_null <- sim_null$metadata
bugs_null <- sim_null$bugs

# Bray–Curtis distance
dist_null <- vegdist(bugs_null, method = "bray")

# Naïve PERMANOVA (ignores subject)
adonis_naive <- adonis2(dist_null ~ exposure, data = meta_null, permutations = 999)

# Repeated-measures PERMANOVA (permute within subject)
perm_rm <- PERMANOVA_repeat_measures(
  formula           = dist_null ~ exposure,
  data              = meta_null,
  blocking_variable = "subject",
  permutations      = 999
)

# Hájek-style repeated-measures test
hajek_rm <- hajek_repeat_measures(
  formula           = dist_null ~ exposure,
  data              = meta_null,
  bugs              = bugs_null,
  blocking_variable = "subject",
  covariate_name    = "exposure",
  permutations      = 999,
  split_ratio       = 0.5,
  numerical_metadata = NULL,
  method            = "bray"
)

adonis_naive
perm_rm
hajek_rm
```

Under the null, all three p-values should be approximately uniform across repeated
simulations; here we see one realization.

# Small simulation study

We now run a **small Monte Carlo experiment** comparing:

- `Naive`: `adonis2(dist ~ exposure)`  
- `RM`: `PERMANOVA_repeat_measures()`  
- `Hajek`: `hajek_repeat_measures()`

under both:

- **null**: `effect_size = 0` (type I error),  
- **alternative**: `effect_size > 0` (power).

To keep the vignette fast, we use small numbers; increase them in your own work.

```{r sim-fun, message=FALSE, warning=FALSE}
run_small_sim <- function(
  n_sim       = 50,
  n_subj      = 20,
  m_per_subj  = 3,
  n_taxa      = 50,
  effect_size = 0
) {
  out <- vector("list", n_sim)

  for (i in seq_len(n_sim)) {
    s <- simulate_microbiome(
      n_subj      = n_subj,
      m_per_subj  = m_per_subj,
      n_taxa      = n_taxa,
      effect_size = effect_size
    )

    meta <- s$metadata
    bugs <- s$bugs
    dist <- vegdist(bugs, method = "bray")

    # Naïve
    a_naive <- adonis2(dist ~ exposure, data = meta, permutations = 499)

    # RM
    a_rm <- PERMANOVA_repeat_measures(
      formula           = dist ~ exposure,
      data              = meta,
      blocking_variable = "subject",
      permutations      = 499
    )

    # Hájek
    a_hajek <- hajek_repeat_measures(
      formula           = dist ~ exposure,
      data              = meta,
      bugs              = bugs,
      blocking_variable = "subject",
      covariate_name    = "exposure",
      permutations      = 499,
      split_ratio       = 0.5,
      numerical_metadata = NULL,
      method            = "bray"
    )

    out[[i]] <- data.frame(
      sim      = i,
      method   = c("Naive", "RM", "Hajek"),
      pval     = c(
        a_naive$`Pr(>F)`[1],
        a_rm$`Pr(>F)`[1],
        a_hajek$pval
      ),
      stringsAsFactors = FALSE
    )
  }

  do.call(rbind, out)
}

set.seed(123)

res_null <- run_small_sim(
  n_sim       = 50,
  effect_size = 0
)

res_alt <- run_small_sim(
  n_sim       = 50,
  effect_size = 1.0  # spike in effect
)

head(res_null)
head(res_alt)
```

## Type I error (null)

```{r null-plots, fig.height=4, fig.width=6}
alpha <- 0.05

# histogram of p-values under the null
ggplot(res_null, aes(x = pval)) +
  geom_histogram(breaks = seq(0, 1, by = 0.05),
                 color = "black", fill = "gray80", closed = "left") +
  facet_wrap(~ method, nrow = 1) +
  geom_hline(yintercept = nrow(res_null) / length(unique(res_null$method)) * alpha,
             colour = "red") +
  theme_bw() +
  labs(x = "P-value", y = "Count",
       title = "Null simulations (effect_size = 0): p-value distributions")

# empirical type I error
res_null %>%
  group_by(method) %>%
  summarise(
    FPR_0.05 = mean(pval < alpha),
    .groups  = "drop"
  )
```

The red line marks the expected number of discoveries at level 0.05 if p-values
were perfectly uniform. In a well-calibrated method, `FPR_0.05` should be near 0.05.

## Power (alternative)

```{r alt-plots, fig.height=4, fig.width=6}
# histogram of p-values under the alternative
ggplot(res_alt, aes(x = pval)) +
  geom_histogram(breaks = seq(0, 1, by = 0.05),
                 color = "black", fill = "gray80", closed = "left") +
  facet_wrap(~ method, nrow = 1) +
  theme_bw() +
  labs(x = "P-value", y = "Count",
       title = "Alternative simulations (effect_size = 1): p-value distributions")

# empirical power at alpha = 0.05
res_alt %>%
  group_by(method) %>%
  summarise(
    Power_0.05 = mean(pval < alpha),
    .groups    = "drop"
  )
```

Under the alternative (`effect_size = 1` here), methods with larger `Power_0.05`
are more sensitive to the spiked microbiome effect.

# Summary

In this vignette we:

- simulated simple repeated-measures microbiome data with subject-level
  heterogeneity and a binary exposure,
- applied:
  - a naïve PERMANOVA that ignores the subject structure,
  - `PERMANOVA_repeat_measures()` using within-subject permutations,
  - `hajek_repeat_measures()` for a Hájek-style repeated-measures test,
- ran a small simulation to compare type I error and power.

You can adapt the simulation settings (number of subjects, samples per subject,
taxa, and effect sizes) to match your study context and to benchmark different
PERMANOVA-style methods using **permabrioche**.

```{r session-info}
sessionInfo()
```


