---
title: "permaBrioche: A Beginner’s Guide to Repeated-Measures PERMANOVA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{permabrioche: beginner repeated-measures workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

# Overview

This vignette explains how to analyze **repeated‑measures microbiome data**
using **permaBrioche**. Specifically, we show:

1.  how to use `PERMANOVA_repeat_measures()`
2.  how to obtain a **null‑centered $$R^2$$**
3.  how to use `hajek_repeat_measures()`
4.  how to decompose the Hajek effect into **location** and **dispersion**

***

# 0. Loading the HMP-based data for this vignette

```{r}
library(permabrioche)
library(vegan)

load(system.file("extdata", "d_hmp.Rda", package = "permabrioche"))    # loads d
load(system.file("extdata", "bugs_hmp.Rda", package = "permabrioche")) # loads bugs

meta <- d              # rename metadata for clarity
bugs <- bugs[, -1]     # drop redundant 'sample' column

# (Good practice: Ensure alignment between metadata and feature matrix)
rownames(meta) <- meta$sample
stopifnot(identical(rownames(meta), rownames(bugs)))

```
***

# 1. The basic data structure

All functions in **permaBrioche** require three aligned objects:

## (a) Sample‑level metadata

A `data.frame` with **one row per sample**, containing at least:

*   `sample` – unique sample ID
*   `subject` – repeated‑measures (blocking) ID
*   `exposure` – a binary covariate of interest

```{r}
head(meta)
```

Each subject may appear in multiple rows.

***

## (b) Feature matrix (`bugs`)

A matrix of features (e.g., taxa abundances):

*   rows = samples
*   columns = features
*   **rownames must match `meta$sample`**

```{r}
dim(bugs)
```

***

## (c) Distance matrix

Use vegdist to get a distance matrix with a chosen method:

```{r}
library(vegan)
dist_bc <- vegdist(bugs, method = "bray")
```

***

# 2. Repeated‑measures PERMANOVA

## Why not `adonis2()`?

Standard PERMANOVA assumes **independent samples**.
In longitudinal or clustered studies, samples from the same subject
may be strongly correlated.

Naively permuting labels across samples can produce false positives (incorrectly low p-values).

***

## `PERMANOVA_repeat_measures()`

This function performs PERMANOVA while:

*   permuting **variant covariates within subjects**, and
*   shuffling **invariant covariates across subjects**.

### Minimal example

```{r}
library(permabrioche)

res_perm <- PERMANOVA_repeat_measures(
  formula           = dist_bc ~ exposure, # adonis2-style formula
  data              = meta,
  sample_id         = "sample",
  blocking_variable = "subject",
  permutations      = 999
)

res_perm
```

***

## How to interpret the output

The output mirrors `adonis2()`:

*   `F` and `Pr(>F)` test whether the exposure is associated
    with multivariate composition.
*   The p‑value is valid **under repeated measures**.

Use this function whenever a subject contributes multiple samples.

***

# 3. Null‑centered $$R^2$$

## Why standard PERMANOVA $$R^2$$ is misleading

As shown in the accompanying manuscript (Effect-Size Estimation and Confounding-Aware Permutations for PERMANOVA), PERMANOVA $$R^2$$ is
**upwardly biased under the null**, especially in small or complex designs.

A value greater than zero does *not* necessarily imply a real effect.

***

## Null‑centered adjustment

`PERMANOVA_repeat_measures()` can subtract the **mean null $$R^2$$**
estimated from permutations.

This yields an effect size interpreted as:

> variation explained **beyond chance under the study design**

***

## Use `center_R2 = TRUE` to compute null‑centered $$R^2$$

```{r}
res_perm_centered <- PERMANOVA_repeat_measures(
  formula           = dist_bc ~ exposure,
  data              = meta,
  sample_id         = "sample",
  blocking_variable = "subject",
  permutations      = 999,
  center_R2         = TRUE
)

res_perm_centered
```

### What to report

*   `R2_centered` – adjusted effect size
*   `Pr(>F)` – same hypothesis test

### Note on Euclidean case

If the distance matrix is computed using Euclidean distance, the null-centering adjustment
has an analytical form. Instead of using the estimation from the above function, it can be simply calculated as $$R^2_{centered} - R^2_{original} - \frac{g-1}{N-1}$$,
where R2_original is the unadjusted estimate (obtained from running adonis2, for example), $$g$$ is the number of levels of the exposure variable (which is 2 for a binary exposure), and $$N$$ is the 
total number of samples.

***

# 4. Hajek distance‑based effect size

## What question does this answer?

PERMANOVA provides a statistically-grounded **test**, but its effect size is somewhat abstract.

The Hajek estimator answers:

> *On average, how far does exposure move samples away from
> a control centroid, in units of the chosen distance metric?*

This new effect size is very cleanly interpretable.

***

## `hajek_repeat_measures()`

The method:

1.  splits subjects into training and test sets
2.  estimates a control centroid using training data
3.  computes distances for test samples
4.  estimates a Hajek average treatment effect
5.  uses subject‑aware permutations for inference

***

### Usage is similar to `PERMANOVA_repeat_measures` except it requires 
specifying the feature matrix: `bugs`, the binary exposure name: `covariate_name`,
and how distances should be computed when constructing the outcome: `method`.

```{r}
res_hajek <- hajek_repeat_measures(
  formula           = dist_bc ~ exposure,
  data              = meta,
  bugs              = bugs,
  sample_id         = "sample",
  blocking_variable = "subject",
  covariate_name    = "exposure",
  permutations      = 999,
  method            = "bray"
)

res_hajek$pval
```

***

## Interpreting the output

*   `observed` – effect size on the distance scale
*   `pval` – permutation p‑value
*   positive value → exposure increases distance from control centroid
*   negative value → exposure decreases distance from control centroid

This effect size has a direct geometric interpretation.

***

# 5. Mean–dispersion decomposition

In the Euclidean case, the Hajek effect can be decomposed into:

*   **location**: change in mean composition
*   **dispersion**: change in variability

***

## Required inputs (important)

The decomposition explicitly requires **correct data levels**:

| Object           | Level            |
| ---------------- | ---------------- |
| `permute_within` | sample           |
| `blocks`         | sample → subject |
| `block_data`     | subject          |
| `bugs`           | sample           |

***

## Example

```{r}
blocks <- meta$subject
names(blocks) <- meta$sample
blocks <- blocks[rownames(bugs)]

permute_within <- data.frame(row.names = rownames(bugs))

block_data <- aggregate(
  exposure ~ subject,
  data = meta,
  FUN = function(x) x[1]
)
row.names(block_data) <- block_data$subject
block_data$subject <- NULL

res_ld <- hajek_repeat_measures_loc_and_disp(
  D              = dist_bc,
  permute_within = permute_within,
  blocks         = blocks,
  block_data     = block_data,
  bugs           = bugs,
  covariate_name = "exposure",
  split_ratio    = 0.5,
  method         = "euclidean"
)

res_ld
```

***

## Interpretation

*   `observed_tau` – total Hajek effect
*   `observed_loc` – mean (centroid) shift
*   `observed_disp` – change in variability

Large location with small dispersion suggests a systematic shift,
while large dispersion suggests increased heterogeneity.

***

# Summary

This vignette demonstrated how to analyze repeated‑measures
microbiome data using **permaBrioche**.

***
