---
title: "permabrioche vignette: HMP-Based Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{permabrioche (HMP-based simulated example)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
  - |
    <style>
      h1.title {
        color: "darkred";
      }
    </style>
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center"
)
set.seed(1)
```

# Overview

This vignette illustrates how to use **permabrioche** for repeated-measures
PERMANOVA on microbiome data derived from the Human Microbiome Project (HMP).

We:

1. load an HMP-based microbiome dataset (`d`, `bugs`) with repeated measures,  
2. construct a **within-subject (variant) varying binary exposure**,  
3. compare:
   - a PERMANOVA implementation that ignores subject,
   - our subject-aware PERMANOVA using `PERMANOVA_repeat_measures()`,
   - our subject-aware Hajek estimator using `hajek_repeat_measures()`,
4. run a **small simulation study** that uses the HMP data as a base:
   - a **null** scenario (no spiked effect of exposure),
   - an **alternative** scenario (spiked microbiome effect).

# Setup and data

```{r setup, message = FALSE, warning = FALSE}
library(permabrioche)
library(dplyr)
library(ggplot2)

set.seed(1)

# -----------------------------------------------------------
# Load HMP-based microbiome data (metadata `d`, counts `bugs`)
# -----------------------------------------------------------
d_path    <- system.file("extdata", "d_hmp.Rda",    package = "permabrioche")
bugs_path <- system.file("extdata", "bugs_hmp.Rda", package = "permabrioche")
obj_d    <- load(d_path)
obj_bugs <- load(bugs_path)
# Explicitly extract first object from each Rda
d    <- get(obj_d[1])
bugs <- get(obj_bugs[1])
rm(obj_d, obj_bugs)

# Drop the unnecessary "sample" column from bugs:
bugs      <- bugs[,-1]

# Keep originals
d_orig    <- d
bugs_orig <- bugs

str(d_orig[, c("subject", "sample")])
dim(bugs_orig)
```

We assume:

- `d` contains at least `subject` (repeated-measures ID) and `sample`,  
- `bugs` is a matrix or data frame of taxa features with rownames equal to
  `d$sample`

# A helper: HMP-based repeated-measures simulator

We now define a simple simulator that:

- selects a subset of subjects and samples from the HMP data,  
- creates a **within-subject (variant) varying binary exposure**,  
- **spikes some taxa** for exposed samples,
- returns metadata, spiked microbiome data, and a Bray–Curtis 'dist' object.

```{r simulator, warning = FALSE}
simulate_hmp_microbiome <- function(
  n_subj           = 30,
  max_samples_subj = 10,
  frac_spike_taxa  = 0.1,
  effect_size      = 0,   # 0 = null, >0 = alternative
  seed             = NULL
) {
  if (!is.null(seed)) set.seed(seed)

  stopifnot(all(c("subject", "sample") %in% colnames(d_orig)))
  stopifnot(identical(rownames(bugs_orig), d_orig$sample))

  # -----------------------------
  # 1. Select subset of subjects and samples
  # -----------------------------
  subj_all <- unique(d_orig$subject)

  if (length(subj_all) < n_subj) {
    warning("Requested n_subj > available; using all subjects.")
    subj_use <- subj_all
  } else {
    subj_use <- sample(subj_all, n_subj, replace = FALSE)
  }

  d_sub <- d_orig[d_orig$subject %in% subj_use, ]

  # Limit to at most max_samples_subj per subject
  d_sub <- d_sub %>%
    group_by(subject) %>%
    slice_head(n = max_samples_subj) %>%
    ungroup()

  bugs_sub <- bugs_orig[match(d_sub$sample, rownames(bugs_orig)), , drop = FALSE]
  stopifnot(identical(rownames(bugs_sub), d_sub$sample))

  rownames(d_sub) <- d_sub$sample

  # -----------------------------
  # 2. Construct within-subject binary exposure
  # -----------------------------
  # For each subject, sample a subject-specific exposure probability p_s,
  # then draw a 0/1 exposure per sample with probability p_s.
  subj_list <- unique(d_sub$subject)
  exposure_vec <- numeric(nrow(d_sub))

  for (s in subj_list) {
    idx <- which(d_sub$subject == s)
    # Subjects are either mostly 0 or mostly 1
    p_s <- sample(c(0.2, 0.8), size = 1)
    exposure_vec[idx] <- rbinom(length(idx), size = 1, prob = p_s)
  }

  d_sub$exposure_num <- exposure_vec
  d_sub$exposure <- factor(exposure_vec, levels = 0:1, labels = c("a", "b"))

  # -----------------------------
  # 3. Spike taxa for the alternative (effect_size > 0)
  # -----------------------------
  bugs_spiked <- as.matrix(bugs_sub)
  n_taxa      <- ncol(bugs_spiked)

  n_spike_taxa <- max(1L, round(frac_spike_taxa * n_taxa))
  spike_taxa   <- sample(colnames(bugs_spiked), n_spike_taxa)

  if (effect_size > 0) {
    # multiplicative spike on spike_taxa for exposed samples
    exposed_idx <- which(d_sub$exposure_num == 1)
    if (length(exposed_idx) > 0) {
      bugs_spiked[exposed_idx, spike_taxa] <-
        bugs_spiked[exposed_idx, spike_taxa] * (1 + effect_size)
    }

    # small subject-level random shift, to mimic subject variation
    for (s in subj_list) {
      idx   <- which(d_sub$subject == s)
      shift <- matrix(runif(n_taxa, min = 0, max = 0.05),
                      nrow = length(idx), ncol = n_taxa, byrow = TRUE)
      bugs_spiked[idx, ] <- bugs_spiked[idx, ] + shift
    }

    # renormalize to compositional rows 
    row_sums <- rowSums(bugs_spiked)
    bugs_spiked <- sweep(bugs_spiked, 1, row_sums, FUN = "/")
    bugs_spiked[is.na(bugs_spiked)] <- 0
  }

  # -----------------------------
  # 4. Bray–Curtis distance
  # -----------------------------
  dist_spiked <- vegdist(bugs_spiked, method = "bray")

  list(
    metadata = d_sub,
    bugs     = bugs_spiked,
    dist     = dist_spiked
  )
}

# Quick check: null data
sim0 <- simulate_hmp_microbiome(effect_size = 0, seed = 1)
str(sim0$metadata)
dim(sim0$bugs)
```

# One-shot example: null scenario (no spiked effect)

We start with a **null** scenario, `effect_size = 0`, where the simulated
exposure has *no* spiked microbiome effect (it still may correlate with the
real underlying HMP data, but we do not add any artificial effect).

```{r one-null, warning = FALSE}
sim_null <- simulate_hmp_microbiome(
  n_subj           = 30,
  max_samples_subj = 10,
  effect_size      = 0,
  seed             = 123
)

meta_null <- sim_null$metadata
bugs_null <- sim_null$bugs
dist_null <- sim_null$dist

# Naïve PERMANOVA (ignores subject)
adonis_naive <- adonis2(dist_null ~ exposure, data = meta_null, permutations = 999)

# Repeated-measures PERMANOVA: permute within subjects
perm_rm <- PERMANOVA_repeat_measures(
  formula           = dist_null ~ exposure,
  data              = meta_null,
  sample_id         = "sample",
  blocking_variable = "subject",
  permutations      = 999
)

# Hájek-style repeated-measures test
hajek_rm <- hajek_repeat_measures(
  formula           = dist_null ~ exposure,
  data              = meta_null,
  sample_id         = "sample",
  bugs              = bugs_null,
  blocking_variable = "subject",
  covariate_name    = "exposure",
  permutations      = 999,
  split_ratio       = 0.5,
  numerical_metadata = NULL,
  method            = "bray"
)

adonis_naive
perm_rm
hajek_rm[names(hajek_rm) != "null_dist"]
```

Under the null, across repeated simulations:

- the vanilla PERMANOVA may show inflated type I error because it ignores
  subject-level confounding,  
- the repeated-measures and Hajek methods should have better-calibrated
  p-values. 
- this is only for one replicate though, so let us take a look at some Monte Carlo samples.

# A small HMP-based simulation study

We now build a small Monte Carlo experiment where each replicate:

1. Simulates an HMP-based dataset with repeated measures,  
2. Fits:
   - `Naive`: `adonis2(dist ~ exposure)`,
   - `RM`: `PERMANOVA_repeat_measures()`,
   - `Hajek`: `hajek_repeat_measures()`,  
3. Stores the p-values for comparison.

We consider:

- a **null** scenario (`effect_size = 0`),  
- an **alternative** scenario (`effect_size > 0`, with multiplicative spikes).

To keep the vignette fast, we use a modest `n_sim`; you can increase it if you would like.

```{r sim-fun, message=FALSE, warning=FALSE}
run_small_sim_hmp <- function(
  n_sim           = 30,
  n_subj          = 30,
  max_samples_subj = 10,
  frac_spike_taxa = 0.1,
  effect_size     = 0
) {
  out <- vector("list", n_sim)

  for (i in seq_len(n_sim)) {

    s <- simulate_hmp_microbiome(
      n_subj           = n_subj,
      max_samples_subj = max_samples_subj,
      frac_spike_taxa  = frac_spike_taxa,
      effect_size      = effect_size,
      seed             = i
    )

    meta <- s$metadata
    bugs <- s$bugs
    dist <- s$dist

    # Naïve PERMANOVA
    a_naive <- adonis2(dist ~ exposure, data = meta, permutations = 499)

    # Repeated-measures PERMANOVA
    a_rm <- PERMANOVA_repeat_measures(
      formula           = dist ~ exposure,
      data              = meta,
      sample_id         = "sample",
      blocking_variable = "subject",
      permutations      = 499
    )

    # Hájek repeated-measures test
    a_hajek <- hajek_repeat_measures(
      formula           = dist ~ exposure,
      data              = meta,
      sample_id         = "sample",
      bugs              = bugs,
      blocking_variable = "subject",
      covariate_name    = "exposure",
      permutations      = 499,
      split_ratio       = 0.5,
      numerical_metadata = NULL,
      method            = "bray"
    )

    out[[i]] <- data.frame(
      sim    = i,
      method = c("Naive", "RM", "Hajek"),
      pval   = c(
        a_naive$`Pr(>F)`[1],
        a_rm$`Pr(>F)`[1],
        a_hajek$pval
      ),
      stringsAsFactors = FALSE
    )
  }

  do.call(rbind, out)
}

set.seed(123)

# Null: no spiked effect from exposure
res_null <- run_small_sim_hmp(
  n_sim           = 3,
  n_subj          = 30,
  max_samples_subj = 10,
  frac_spike_taxa = 0.1,
  effect_size     = 0
)

# Alternative: clear spiked effect on a subset of taxa
res_alt <- run_small_sim_hmp(
  n_sim           = 3,
  n_subj          = 30,
  max_samples_subj = 10,
  frac_spike_taxa = 0.1,
  effect_size     = 25    # multiplicative spike
)

head(res_null)
head(res_alt)
```

## Type I error under the null

We first visualize p-values when there is **no spiked effect** (`effect_size = 0`).

```{r null-plots, fig.height=4, fig.width=6}
alpha <- 0.05

n_per_method <- nrow(res_null) / length(unique(res_null$method))

ggplot(res_null, aes(x = pval)) +
  geom_histogram(
    breaks = seq(0, 1, by = 0.05),
    color  = "black",
    fill   = "gray80",
    closed = "left"
  ) +
  geom_hline(
    yintercept = n_per_method * alpha,
    colour     = "red",
    linewidth  = 1
  ) +
  facet_wrap(~ method, nrow = 1) +
  theme_bw() +
  labs(
    title = "Null simulations (HMP-based, effect_size = 0): p-value distributions",
    x     = "P-value",
    y     = "Count"
  )

# Empirical type I error at alpha = 0.05
res_null %>%
  group_by(method) %>%
  summarise(
    FPR_0.05 = mean(pval < alpha),
    .groups  = "drop"
  )
```

The red line shows the expected bin count under a perfect Uniform(0,1)
distribution at significance level 0.05. A well-calibrated method will have
`FPR_0.05` close to 0.05.

## Power under the alternative

We now look at the **alternative** scenario, where the exposure is artificially
linked to a spike in microbiome composition.

```{r alt-plots, fig.height=4, fig.width=6}
alpha <- 0.05

n_per_method_alt <- nrow(res_alt) / length(unique(res_alt$method))

ggplot(res_alt, aes(x = pval)) +
  geom_histogram(
    breaks = seq(0, 1, by = 0.05),
    color  = "black",
    fill   = "gray80",
    closed = "left"
  ) +
  geom_hline(
    yintercept = n_per_method_alt * alpha,
    colour     = "red",
    linewidth  = 1
  ) +
  facet_wrap(~ method, nrow = 1) +
  theme_bw() +
  labs(
    title    = "Alternative simulations (HMP-based, effect_size = 5): p-value distributions",
    subtitle = "Red line is the null-reference count at alpha = 0.05",
    x        = "P-value",
    y        = "Count"
  )

# Empirical power at alpha = 0.05
res_alt %>%
  group_by(method) %>%
  summarise(
    Power_0.05 = mean(pval < alpha),
    .groups    = "drop"
  )
```

Under a strong enough spiked effect, well-powered methods will concentrate p-values
near 0 and show larger `Power_0.05`.

# Summary

In this vignette we:

- started from real **HMP-based microbiome data** (`d`, `bugs`) with repeated measures,  
- defined a simple HMP-based **simulator** that:
  - selects a subset of subjects and samples,
  - constructs a **within-subject binary exposure**,
  - **spikes some taxa** to form an alternative hypothesis,  
- applied:
  - a naive PERMANOVA (`adonis2(dist ~ exposure)`),
  - `PERMANOVA_repeat_measures()` (within-subject permutations),
  - `hajek_repeat_measures()` (Hajek-style repeated-measures test),  
- ran a **small Monte Carlo study** to inspect:
  - type I error (null, `effect_size = 0`),
  - power (alternative, `effect_size > 0`).

